D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-29 ootavacuous.tex
Add Stephen Dolan example to the undefined-behavior appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@

However, some kernels and applications have ways of handling even
uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one user process, and might react by
killing that user process.
Some user applications might be note that the affected data affected only
a particular computation, and might react by restarting that computation.
In both cases, it is only reasonable to react to an uncorrectable ECC
error in a load from a read-only mapped file by re-reading that data
from that file, then restarting that load instruction.
In all cases, the kernel and the user application would need to continue
execution, and would thus require an exact exception.

@@

ECC errors are detected near the end of their corresponding load
instructions' execution, so that any need for exact exceptions rules
out the commiting of any subsequent stores before the load has fetched
its value.

@@

\begin{enumerate}
\item	Some hardware might choose to forego ECC, for example, in order to
	reduce cost or to improve energy efficiency.
\item	ECC error correction might still be done in hardware, avoiding
	the need for precise software-visible exceptions.
\item	Some systems might prefer to immediately shut down, forbidding
	software from continuing after an uncorrectable ECC error,
	perhaps due to safety considerations.
\item	Someone might come up with a clever way of correcting ECC errors
	in firmware, avoiding the need for precise software-visible
	exceptions.
\item	As late as early 2024, some GPUs have been observed reordering
	earlier loads against later stores.
	Other hardware architects might therefore feel the need to keep
	this option open for their own systems.
\end{enumerate}

>>>>

\subsection{Lifetime-Based Undefined Behavior}
\label{sec:Lifetime-Based Undefined Behavior}

Not all undefined behavior can be backstopped using
\co{std::unreachable()}.

\begin{listing}[tbp]
\begin{verbatim}
 1 int *foo(int *r)
 2 {
 3   b.store(r, memory_order_relaxed);
 4   return &global;
 5 }
 6
 7 int go(int *p)
 8 {
 9   int r;
10   int *q = foo(&r);
11
12   r = 2;
13   *p = 1;
14   *q = r;
15   return *q - r;
16 }
17
18 void t1(void)
19 {
20   int r1 = go(a.load(memory_order_relaxed));
21   printf("go() = %d\n", r1);
22 }
23
24 void t2(void)
25 {
26   int *r2 = b.load(memory_order_relaxed);
27   a.store(r2, memory_order_relaxed);
28 }
\end{verbatim}
\caption{Lifetime-Based Undefined Behavior}
\label{lst:Lifetime-Based Undefined Behavior}
\end{listing}

Listing~\ref{lst:Lifetime-Based Undefined Behavior}
shows an example loosely based on one presented by Stephen Dolan\footnote{
	\url{https://stedolan.net/talks/2024/loadstore}}
in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
local variable \co{r}.
However, \co{foo()} uses a relaxed store to place this pointer in
an atomic variable \co{b}.
Because this store is relaxed, it is not ordered against other code
in \co{go()} and \co{t1()}, which means that \co{t2()} might well
read this pointer from \co{b} either before the beginning or after
the end of \co{r}'s lifetime, neither of which is well-defined.

Of course, any attempt to dereference the pointer read from either
\co{a} or \co{b} could result in undefined behavior.

The lesson is clear:
Properly synchronize reads and writes of pointers with the lifetime
of the pointed-to object.

This example could make lines~3 and~27 be \co{memory_order_release} and
lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
a pointer to \co{r} before the lifetime of \co{r} has started.
Note that \co{std::unreachable()} does not suffice because the CPU
can also reorder relaxed stores.
Note also that load-store ordering is not sufficient to prevent a
pointer to \co{r} from being stored to \co{b} prior to the start of
\co{r}'s lifetime because there is no atomic load between the start
of that lifetime and the relaxed store to \co{b}.

There are a number of ways to prevent carrying a pointer to \co{r} after
the lifetime of \co{r} has ended, and these are left as an exercise for
the reader.

@@

However, some kernels and applications have ways of handling even
uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one user process, and might react by
killing that user process, taking care to account for memory shared with
other processes.
Some user applications might be note that the affected data affected only
a particular computation, and might react by restarting that computation.
In both cases, it is only reasonable to react to an uncorrectable ECC
error in a load from a read-only mapped file by re-reading that data
from that file, then restarting that load instruction.
In all cases, the kernel and the user application would need to continue
execution, and would thus require an exact exception.

@@

ECC errors are detected near the end of their corresponding load
instructions' execution, so that any need for exact exceptions rules out
the commiting of any subsequent stores until after the load has fetched
(and possibly corrected) its value.

@@

\begin{enumerate}
\item	Some hardware might choose to forego ECC, for example, in order to
	reduce cost for low-end systems or to improve energy efficiency
	for battery-powered systems.
\item	ECC error correction might still be done in hardware, avoiding
	the need for precise software-visible exceptions.
\item	Some systems might prefer to immediately shut down, forbidding
	software from continuing after an uncorrectable ECC error,
	perhaps due to safety considerations.
\item	Someone might come up with a clever way of correcting ECC errors
	in firmware, avoiding the need for precise software-visible
	exceptions.
\item	As late as early 2024, some GPGPUs have been observed reordering
	earlier loads against later stores.
	Other hardware architects might therefore feel the need to keep
	this option open for their own systems.
\end{enumerate}
